---
author: Staycee (Abby) Surmeier
title: "practice_final"
output: html_document
---

```{r echo = TRUE, warning = FALSE, cache = TRUE}

#  Read in data
data <- read.csv("titanic.csv", na.strings =  c('','?'))
str(data)
summary(data)
head(data)
dim(data)

#de-identify data
data <- data[,-4]
# 891 obs. of  12 variables
```
General observations/questions:
- mostly quant data --> predict which passengers died on the titanic?
- features mostly quantitative 
  -linear relationships?
  -don't know much --> will need to visualize and perhaps cluster? KNN?
- binary classification (survival) --> change to factor
  - KNN
  - log regression
  - random forest
  
```{r echo = TRUE, warning = FALSE, cache = TRUE}

data$Survived <- as.factor(data$Survived)
```

  
```{r echo = TRUE, warning = FALSE, cache = TRUE}
#Missing data
sum(is.na(data))
missing <- as.table(sapply(data, function(x) sum(is.na(x))))
library(naniar)
library(ggplot2)
miss_plot <- gg_miss_var(data)
miss_plot
sum(is.na(data$Cabin))
sum(is.na(data$Age))

```

There are 866 values missing, 687 (79%) in the Cabin column and 177(20%) in the age column. Will ignore these observations in further observations. 

##Importance of columns with missing values
 

library(ggplot2)
cab_survived <- ggplot(data, aes(x = cabin))
```{r echo = TRUE, warning = FALSE, cache = TRUE}
data <- data[,-10]
dim(data)
sum(is.na(data))

```

Now working with 891  observations of 10 variables. 177 NA values remain in age column. Will run KNN then impute if accuracy is poor. 

```{r echo = TRUE, warning = FALSE, cache = TRUE}
levels <- levels(data$Cabin)
levels[length(levels) + 1] <- "None"

# refactor Cabin to include "None" as a factor level
# and replace NA with "None"
data$Cabin <- factor(data$Cabin, levels = levels)
data$Cabin[is.na(data$Cabin)] <- "None"

```

# Get levels and add "None" for missing

```{r echo = TRUE, warning = FALSE, cache = TRUE}

```


```{r echo = TRUE, warning = FALSE, cache = TRUE}

```

```{r echo = TRUE, warning = FALSE, cache = TRUE}

```
